---
title: "Correlation is not Causation"
subtitle: "Week 5 Lecture 2"
format: 
  live-revealjs:
    slide-number: true
    show-slide-number: all
    incremental: true
engine: knitr
resources:
  - happiness_polity_2018.csv
---

{{< include ../_extensions/r-wasm/live/_knitr.qmd >}}

## Tricky parts of the scientific method

- Coming up with a good theory in the first place
- Coming up with appropriate hypotheses to evaluate these theories
- Coming up with empirical tests that effectively test whether the hypothesis holds or not

:::{.fragment}
This last step is much trickier than it looks! Beware **observational equivalence:** *when evidence that could be consistent with a hypothesis could also be interpreted as inconsistent due to alternative explanations.*
:::

## A theory of happiness and longevity

- **Theory:** happiness causes a longer life.
- **Hypothesis:** happier countries have people who live longer.
- **Test:** analysis of `happiness_data`

## Back to our happiness dataset

```{webr}
#| edit: false
#| autorun: false
#| echo: true
#| message: false
#| warning: false
library(ggplot2)
happiness_data <- read.csv("happiness_polity_2018.csv")
head(happiness_data)
```

## Codebook

```{r}
library(knitr)
library(kableExtra)
library(ggplot2)
happiness_data <- read.csv("happiness_polity_2018.csv")

vars <- data.frame(
  Variable = c(
    "polity2", "polity2_cat", "gdpcapita", "gdpcapita_cat",
    "happiness", "happiness_cat", "life_expectancy", "life_expectancy_cat"
  ),
  Description = c(
    "Polity Score: 21-point scale from -10 (hereditary monarchy) to +10 (consolidated democracy)",
    "Political category: autocracies, anocracies (semi-democracies), or democracies",
    "GDP per capita (economic output per person)",
    "Income category based on GDP per capita",
    "Happiness index from surveys on quality-of-life factors",
    "Happiness category based on the happiness index",
    "Average life expectancy, in years",
    "Life expectancy category based on the life expectancy value"
  )
)

kable(vars, format = "html", col.names = c("Variable", "Description")) %>%
  kable_styling(font_size = 18, full_width = FALSE) %>%
  column_spec(1, monospace = TRUE, width = "3cm") %>%
  column_spec(2, width = "9cm")
```

## Do happier countries have people who live longer?

```{webr}
# let's make a ggplot using happiness and life_expectancy from happiness_data.
```

## So does being happy *cause* people to live longer?

- We would have to assume happiness $\rightarrow$ longevity
- But what if it's actually longevity $\rightarrow$ happiness?
  - **Reverse causation:** the possibility that instead of A causing B, B causes A
  
:::{.fragment}
**Correlation is not causation!**
:::
  
## Another possibility
  
```{r}
library(dplyr)
library(tidyr)
happiness_data %>%
  pivot_longer(c(happiness, life_expectancy), names_to = "dv", values_to = "value") %>%
  ggplot(aes(x = polity2, y = value)) +
  geom_point() +
  geom_smooth(method = "lm", level = F) +
  facet_wrap(~dv, scales = "free_y", strip.position = "left") +
  theme(strip.placement = "outside",
        strip.background = element_blank(),
        strip.text = element_text(size = 12)) +
  ylab("")
```

## Omitted variable bias

- Instead of A $\rightarrow$ B, C $\rightarrow$ A and B
- Even though A doesn't cause B, A will be related to B
- What other omitted variables can we imagine in this case?

:::{.fragment}
**Correlation is not causation!**
:::

---

*One of the most important skills you will learn in this class is how to tell stories about why a correlation isn't necessarily causal.*

## So can we test our original hypothesis with these data?

```{r}
ggplot(happiness_data, aes(x = happiness, y = life_expectancy)) +
  geom_point() +
  geom_smooth(method = "lm", level = F)
```

## Summing up

- **Original theory:** A $\rightarrow$ B
- **Reverse causation:** B $\rightarrow$ A
- **Omitted variable bias:** C $\rightarrow$ both A and B

